---
title: "r workshop"
author: "David Costalago"
date: "21 de enero de 2019"
output:
  pdf_document: default
  html_document: default
---

### Creating objects in r
```{r setup, warning = FALSE}
library(tinytex)
```

# Day 1

```{r arithmetic}
3 + 5

weight_kg <- 55
```

## Vectors and data types

This section describes some basic data types in r

```{r}

weight_g <- c(50,60,65,82)

animals <- c("mouse", "rat", "dog")

```

### Data Frames

Next we look at the structure of Data Frames

```{r}
library(tidyverse)

#utils::download.file() #this forces r to use the donwload.file function within the package ´utils´

download.file(url="https://ndownloader.figshare.com/files/2292169", destfile = "read_data/portal_data_joined.csv")

library(here) # This package makes working directories and file paths easy

surveys <- read_csv(here("read_data", "portal_data_joined.csv"))

surveys

dim(surveys)

summary(surveys)
```

### Indexing and subsetting data frames

First lets use square bracket subsetting.


```{r}

# First define the row coordinate, and then the column. 
surveys[1,1]

# Defining only which column we want  will return a data frame
surveys[1]

surveys[1:3, 7] #gives rows 1 to 3 in column 7

surveys[,-7] #all the rows and columns except column 7
surveys[, -c(1:5)]
```

## Data manipulation
Square brackets are good when the coordinates of your data frame are fixed. But what happens if the coordinates change (e.g. you created new columns)

Key functions for data manipulation in dplyr:

* select(): subsetting columns
* filter(): subsets of rows based on condiitons
* mutate(): create new columns, nased on information from other columns
* group_by(): creates groups based on categorical data in a column
* summarize(): creates summary stats on grouped data
* arrange(): sort results
* count(): gives a count of discrete values

```{r}
select(surveys, plot_id, species_id, weight)

#Negative subsetting
select(surveys, -record_id)


filter(surveys, year == 1995,
       species_id == "NL")
```

## Pipes

Pipes allow you yo chain together dplyr functions.

%>% or ctrl-shift-m

```{r}
# Write multiple arguments in a sentence using pipes
surveys %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)

surveys_sml <- surveys %>% 
   filter(weight < 5) %>% 
  select(species_id, sex, weight)

```


* Challenge #1

Using pipes, subset the surveys dataframe to include animals collected before 1995 and retain only the columns year, sex and weight.

```{r challenge 1}

surveys_animal_pre1995 <- surveys %>% 
   filter(year < 1995) %>% 
  select(year, sex, weight)

```

```{r}
surveys %>% 
  mutate(weight_kg = weight / 1000,
         weight_kg2 = weight_kg * 2)

surveys <- surveys %>% 
  drop_na(weight) %>% 
  mutate(mean_weight = mean(weight))

mean(surveys$weight)
```


* Challenge #2

Contains only the species_id column, has a new column called hindfoot_half that are half the hindfoot_length values. Also, in the new hinfoot_half column there are no NAs and values are less than 30

```{r}
surveys_new <- surveys %>% 
  mutate(hindfoot_half = hindfoot_length / 2) %>% 
  drop_na(hindfoot_half) %>% 
  filter(hindfoot_half < 30) %>% 
  select(species_id)


```


```{r}
surveys %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE)) %>% 
  arrange(min_weight) # arrange is done by ascending order by default

```

```{r}
surveys %>% 
  count(sex) #handy to get samples sizes for different groups

# the above code is synonomous
surveys %>%  
  group_by(sex) %>% 
  summarize(count = n ())
```

* Challenge #3

1. How many animals were caught in each plot_type surveyed.

2. Use group_by and summarize to find the mean, min and max of hindfoot length (using species_id) for each species. Also, add the number of obervacions (hint: see ?n).

3. What was the heaviest animal measured in each year? Return the columns year, genus, species_id, and weight

```{r}
#1
surveys %>% 
  count(plot_type)

#2
surveys %>%  
  group_by(species_id) %>% 
  summarize(count = n (),
            mean(hindfoot_length, na.rm = TRUE),
            max(hindfoot_length, na.rm = TRUE),
            min(hindfoot_length, na.rm = TRUE))

#3
surveys %>% 
  group_by(year, genus, species_id) %>% 
  summarize(max_weight = max(weight, na.rm = TRUE)) %>% 
  arrange(desc(max_weight)) 

max_weights <- surveys %>% 
  drop_na(weight) %>% 
  group_by(year) %>% 
  filter(weight == max(weight)) %>% 
  select(year, genus, species, weight) %>% 
  arrange(year) %>% 
  unique() 
```

## Export our data
```{r}
write_csv(max_weights, here("write_data", "max_weights.csv"))
```

# Day 2

## Tidy data in Spreadsheets

Today we look at tidying data

The functions we use for tidying data are:

* tidyr::spread()
* tidyr::gather()

### spread()

spread() takes 3 principal arguments:
1. the data
2. the _key_ column variable will become the new column names
3. the value column variable, which will fill the new column variables

We're going to use the surveys dataset

```{r}
library(readr)
surveys <- read_csv(here("read_data", "surveys.csv"))

# Create a wide data format of surveys using spread()

surveys_gw <- surveys %>% 
  drop_na(weight) %>% 
  group_by(species_id) %>% 
  summarize(mean_weight = mean(weight))

str(surveys_gw)

# Now we want to sprread te data set out into a wide format

wide_surveys_gw <- surveys_gw %>% 
  spread(key = species_id, value = mean_weight)

# Now, using gather() we are going back to long from the wide data set we just created.

```

### gather

gather() takes 4 arguments:
1. data
2. key
3. value
4. names of columns we use to fill the key variable (or drop)

```{r}
long_surveys <- wide_surveys_gw %>% 
  gather(key = species_id, value = mean_weight)
```


## Sending Tidy Data

### Changelog

* Update your changelog with changes you make to your raw data, or other significant changes or additions to your projects!

### Data Dictionary


```{r}
tigy_gsi <- read_csv(here("write_data", "tidy_gsi.csv"))
```

